---
title: "Data Integration Report: RB5 & Biovolume Dataset"
author: "Ariel Koh"
date: "21/04/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

## Overview
This document compiles and cleans several datasets related to RB5 monitoring, biovolume measurements, predator densities, water levels, temperature, pH, and dissolved oxygen (DO) levels for the period 2004–2016. The goal is to merge all data into a unified format suitable for analysis.

## Load and Clean Datasets

### 1. Monitoring Data
Loads RB5 monitoring data, filters for the target period, and drops unused variables.
```{r}
RB5_monitoring_data <- read.csv("../data/csv/RB5_monitoring_data_monthly_average_1980-2023.csv")
RB5_monitoring_data <- RB5_monitoring_data %>%
  filter(between(year, 2004, 2016)) %>%
  select(-RB5.Cond, -RB5.NO2, -RB5.NH4, -RB5.PP, -RB5.PSiO2, -RB5.SURP, -RB5.TSi, -RB5.TSP, -RB5.TON) %>%
  mutate(YearMonth = as.Date(YearMonth))
```

### 2. Biovolume Data
Processes monthly biovolume values and standardizes date format.
```{r}
biovol_fulldata <- read.csv("../data/csv/biovolfulldata_2004-2016.csv") %>%
  mutate(YearMonth = as.Date(YearMonth))
```

### 3. Predator (Daphnia) Data
Selects and processes zooplankton data, aggregates monthly means, and filters for the target timeframe.
```{r}
predator_data <- read.csv("../data/csv/crustaceanzooplankton_1980-2023.csv") %>%
  select(Date, Daphnia.hyalina.ind.L) %>%
  mutate(Date = dmy(Date), year = year(Date), month = month(Date)) %>%
  filter(!is.na(Date)) %>%
  mutate(Daphnia = as.numeric(Daphnia.hyalina.ind.L)) %>%
  select(-Daphnia.hyalina.ind.L) %>%
  group_by(year, month) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE), .groups = 'drop') %>%
  mutate(YearMonth = as.Date(paste(year, month, "01", sep = "-"))) %>%
  filter(between(YearMonth, as.Date("2004-01-01"), as.Date("2016-12-01")))
```

### 4. Water Level Data
Combines two water level datasets and computes monthly means.
```{r}
waterlevelp2 <- read.csv("../data/csv/waterlevels_1993-2007.csv") %>%
  rename(waterlevel.masl = Harbour.masl) %>%
  mutate(Date = dmy(Date)) %>%
  select(Date, waterlevel.masl)

waterlevelp3 <- read.csv("../data/csv/waterlevels_2008-2013.csv") %>%
  rename(Date = SAMPLE_DATE, waterlevel.masl = Level.maod) %>%
  mutate(Date = dmy(Date)) %>%
  select(Date, waterlevel.masl)

fullwaterlevel <- bind_rows(waterlevelp2, waterlevelp3) %>%
  mutate(year = year(Date), month = month(Date)) %>%
  group_by(year, month) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE), .groups = 'drop') %>%
  mutate(YearMonth = as.Date(paste(year, month, "01", sep = "-"))) %>%
  filter(between(year, 2004, 2016))
```

### 5. Temperature and SD Data
Processes temperature and standard deviation data, summarising to monthly means.
```{r}
tempsddata <- read.csv("../data/csv/Temp_SD_2004-2016.csv") %>%
  mutate(Date = dmy(Date), RB5.SD = as.numeric(RB5.SD), RB5.Temp = as.numeric(RB5.Temp),
         year = year(Date), month = month(Date)) %>%
  group_by(year, month) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE), .groups = 'drop') %>%
  mutate(YearMonth = as.Date(paste(year, month, "01", sep = "-")))
```

### 6. pH Data (2010–2016)
Processes pH values, available only for 2010–2016, into monthly means.
```{r}
pHdata <- read.csv("../data/csv/pH_2010-2016.csv") %>%
  select(Date, Value) %>%
  mutate(Date = dmy(Date), RB5.pH = Value,
         year = year(Date), month = month(Date)) %>%
  group_by(year, month) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE), .groups = 'drop') %>%
  mutate(YearMonth = as.Date(paste(year, month, "01", sep = "-")))
```

### 7. DO Data (2010–2016)
Filters and processes dissolved oxygen (% saturation) into monthly averages.
```{r}
DOdata <- read.csv("../data/csv/DO_2010-2016.csv") %>%
  filter(Units == "%") %>%
  select(Date, Value) %>%
  mutate(Date = dmy(Date), RB5.DO = Value,
         year = year(Date), month = month(Date)) %>%
  group_by(year, month) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE), .groups = 'drop') %>%
  mutate(YearMonth = as.Date(paste(year, month, "01", sep = "-")))
```

## Merge All Data
This step incrementally merges all cleaned datasets into one by matching on the `YearMonth` column.
```{r}
all_data1 <- full_join(RB5_monitoring_data, select(predator_data, YearMonth, Daphnia), by = "YearMonth")
all_data2 <- full_join(biovol_fulldata, select(fullwaterlevel, YearMonth, waterlevel.masl), by = "YearMonth")
all_data <- full_join(all_data2, all_data1, by = "YearMonth")

all_data <- all_data %>%
  left_join(tempsddata, by = "YearMonth") %>%
  mutate(RB5.SD = coalesce(RB5.SD.x, RB5.SD.y),
         RB5.Temp = coalesce(RB5.Temp.x, RB5.Temp.y)) %>%
  select(-RB5.SD.x, -RB5.SD.y, -RB5.Temp.x, -RB5.Temp.y) %>%
  left_join(pHdata, by = "YearMonth") %>%
  mutate(RB5.pH = coalesce(RB5.pH.x, RB5.pH.y)) %>%
  select(-RB5.pH.x, -RB5.pH.y) %>%
  left_join(DOdata, by = "YearMonth") %>%
  mutate(RB5.DO = coalesce(RB5.DO.x, RB5.DO.y)) %>%
  select(-RB5.DO.x, -RB5.DO.y)
```

## Final Cleaning
Refines merged dataset by selecting relevant columns, deriving seasonal categories, and calculating total biovolume.
```{r}
all_data <- all_data %>%
  distinct(YearMonth, .keep_all = TRUE) %>%
  select(YearMonth, year.y, month.y, Cryto.Biovolume, Cyano.Biovolume,
         Diatoms.Biovolume, Greens.Biovolume, waterlevel.masl, Daphnia,
         starts_with("RB5")) %>%
  rename(Year = year.y, Month = month.y) %>%
  mutate(Season = case_when(
    Month %in% c(12, 1, 2) ~ "Winter",
    Month %in% c(3, 4, 5) ~ "Spring",
    Month %in% c(6, 7, 8) ~ "Summer",
    Month %in% c(9, 10, 11) ~ "Autumn"
  )) %>%
  mutate(Total.Biovolume = Cryto.Biovolume + Cyano.Biovolume +
                           Diatoms.Biovolume + Greens.Biovolume) %>%
  relocate(Season, .after = Month) %>%
  relocate(Total.Biovolume, .after = Greens.Biovolume)
```

## Save Final Dataset
Exports the cleaned and merged dataset as CSV files for further analysis.
```{r}
write.csv(all_data, "../data/csv/alldata-2004-2016.csv", row.names = FALSE)
write.csv(all_data, "../data/csv/alldata-2004-2016v2.csv", row.names = FALSE)
```

## Conclusion
All datasets have been merged and cleaned into a consistent structure for further analysis. The final dataset contains monthly observations from 2004 to 2016, with environmental, biological, and chemical measurements aligned by date.

