---
title: "Random Forest Regression for Predicting Total Biovolume"
author: "Ariel Koh"
date: "22/04/2025"
output:
  word_document:
    toc: true
    toc_depth: 2
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
library(tidyverse)
library(caret)
library(randomForest)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(888)
```

## Load and Prepare Data

```{r load-data}
alldata <- read.csv("../data/csv/alldata-2004-2016v2.csv") %>%
  filter(!is.na(Total.Biovolume)) %>%
  select(-YearMonth, -Month, -Cryto.Biovolume, -Cyano.Biovolume, -Diatoms.Biovolume, -Greens.Biovolume, -RB5.ChlA) %>%
  mutate(Year = as.numeric(Year), Season = as.factor(Season))
summary(alldata)
```

## Random Forest with Imputed Data

### Impute Missing Values
```{r impute}
data_impute <- rfImpute(Total.Biovolume ~ ., data = alldata, ntree = 500)
```

### Train-Test Split
```{r split1}
trainIndex1 <- createDataPartition(data_impute$Total.Biovolume, p = 0.7, list = FALSE)
trainData1 <- data_impute[trainIndex1, ]
testData1 <- data_impute[-trainIndex1, ]
```

### Fit Random Forest
```{r rf1}
rfmodel1 <- randomForest(Total.Biovolume ~ ., data = trainData1, ntree = 500, mtry = sqrt(ncol(trainData1)-1), importance = TRUE)
print(rfmodel1)
rfmodel1$importance
varImpPlot(rfmodel1)
```

### Evaluate Model
```{r eval1}
predictions <- predict(rfmodel1, testData1)
rmse <- sqrt(mean((predictions - testData1$Total.Biovolume)^2))
rsq <- cor(predictions, testData1$Total.Biovolume)^2
cat("R²:", rsq, "\nRMSE:", rmse)
```

## Tune Model Hyperparameters

```{r tune1}
tuned_model <- tuneRF(trainData1[, -which(names(trainData1) == "Total.Biovolume")], trainData1$Total.Biovolume, stepFactor = 1.5, improve = 0.01, ntreeTry = 500, trace = TRUE)
```

### Final Tuned Model
```{r rf1opt}
rfmodel1opt <- randomForest(Total.Biovolume ~ ., data = trainData1, ntree = 500, mtry = 6, importance = TRUE)
print(rfmodel1opt)
rfmodel1opt$importance
predictions1opt <- predict(rfmodel1opt, testData1)
rmse1opt <- sqrt(mean((predictions1opt - testData1$Total.Biovolume)^2))
rsq1opt <- cor(predictions1opt, testData1$Total.Biovolume)^2
cat("R²:", rsq1opt, "\nRMSE:", rmse1opt)
varImpPlot(rfmodel1opt)
```

## Random Forest on Complete Cases Only

```{r no-na}
set.seed(40)
noNAdata <- alldata %>% drop_na()
```

### Train-Test Split
```{r split2}
trainIndex2 <- createDataPartition(noNAdata$Total.Biovolume, p = 0.7, list = FALSE)
trainData2 <- noNAdata[trainIndex2, ]
testData2 <- noNAdata[-trainIndex2, ]
```

### Fit Model
```{r rf2}
rfmodel2 <- randomForest(Total.Biovolume ~ ., data = trainData2, ntree = 500, mtry = sqrt(ncol(trainData2)-1), importance = TRUE)
print(rfmodel2)
rfmodel2$importance
predictions2 <- predict(rfmodel2, testData2)
rmse2 <- sqrt(mean((predictions2 - testData2$Total.Biovolume)^2))
rsq2 <- cor(predictions2, testData2$Total.Biovolume)^2
cat("R²:", rsq2, "\nRMSE:", rmse2)
varImpPlot(rfmodel2)
```

### Tune Final Model
```{r tune2}
tuned_model2 <- tuneRF(trainData2[, -which(names(trainData2) == "Total.Biovolume")], trainData2$Total.Biovolume, stepFactor = 1.5, improve = 0.01, ntreeTry = 500, trace = TRUE)
```

```{r rf2opt}
rfmodel2opt <- randomForest(Total.Biovolume ~ ., data = trainData2, ntree = 500, mtry = 4, importance = TRUE)
print(rfmodel2opt)
rfmodel2opt$importance
predictions2opt <- predict(rfmodel2opt, testData2)
rmse2opt <- sqrt(mean((predictions2opt - testData2$Total.Biovolume)^2))
rsq2opt <- cor(predictions2opt, testData2$Total.Biovolume)^2
cat("R²:", rsq2opt, "\nRMSE:", rmse2opt)
```

### Variable Importance Plot (Renamed)
```{r importance-plot}
imp_df <- as.data.frame(importance(rfmodel2opt, type = 1))
imp_df$vars <- rownames(imp_df)
imp_df$vars <- recode(imp_df$vars,
                      "RB5.SRP" = "SRP",
                      "RB5.NO3" = "Nitrate",
                      "RB5.SRSi" = "SRSi",
                      "RB5.Temp" = "Temperature",
                      "waterlevel.masl" = "Water Level",
                      "RB5.DO" = "DO",
                      "RB5.pH" = "pH",
                      "RB5.TP" = "TP",
                      "RB5.SD" = "Secchi Depth")
imp_df <- imp_df[order(imp_df$`%IncMSE`), ]
dotchart(imp_df$`%IncMSE`, labels = imp_df$vars, xlim = c(-2, max(imp_df$`%IncMSE`, na.rm = TRUE) * 1.1),
         pch = 1, xlab = "% Increase in MSE", ylab = "Predictors")
```

## Conclusion
This analysis used random forest regression to identify key environmental predictors of total phytoplankton biovolume. Both imputed and complete-case approaches yielded consistent variable importance rankings, offering confidence in the robustness of the model outputs.

